|

### 2023-10-27 Paper/draft.Rmd
draft edits

---
title: "Learning with Mental Models: the case of overestimation"  
date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Jimena Galindo"
abstract: "I design a framework and a laboratory experiment that allow for the comparison of multiple theories of misspecified learning. I focus on a framework with endogenous information and a data-generating process ruled by two fundamentals: an ego-relevant parameter and a state. Within this framework I study three forces that can lead to misspecified beliefs: initial misspecifications, learning traps and biased updating. I find that biased updating is the main driver of misspecified beliefs in the lab. In addition, I vary the degree of ego-relevance of the parameter by introducing a stereotype treatment. The data is consistent with biased updating in both cases but for potentially different reasons: when learning about themselves, subjects attribute successes to their own ability and failures to luck. Instead, in the stereotype treatment, they compensate for initial negative biases by over-attributing positive signals to the ability of others. This translates into similar observed choices but different dynamics in beliefs."

@@

bibliography: references.bib
fontsize: 12pt 
geometry: margin=2.5cm

@@

As an example consider a student who needs to decide how much effort to put into 
studying for an exam. Their decision will depend on two factors: their belief about their intrinsic 
ability, and their belief about how difficult the exam will be. The outcome 
they observe will be affected by how much they decide to study. Imagine that the student 
puts in a moderate amount of effort and gets a surprisingly good grade. Did they get a good grade because 
they are smarter than they thought? Or because the exam was easier than they had anticipated? 
Their future exam preparation strategy will depend fundamentally on which line of reasoning they 
take. This feedback loop is referred to as an \emph{endogenous information 
process} and is at the center of the forces I study.

To understand what the main forces at play are, I compare a set of theories that model learning in settings with 
endogenous information, and which can rationalize the persistence of misspecified beliefs. I develop a unifying framework that nests multiple 
theories of learning and generates testable predictions for each of them. 
Then I cast this framework in a laboratory experiment and test the predictions to identify which 
of the theories are consistent with the behavior I observe in the lab.
The experiment features an agent who needs to 
learn two parameters: one that pertains their own characteristics 
(an \emph{ego-relevant} parameter), 
and an exogenously determined state---in the context of the student preparing for exams, 
the two parameters  would be their intrinsic ability and the difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred to as overconfidence 
(or underconfidence) and 
have been documented by behavioral scientists\footnote{\citet{kelley1980} 
provides a review of the psychology literature.} and economists\footnote{See \citet{benjamin2019} 
for a review of the literature in economics.} in a variety of settings.  @Oster2013 find that 
subjects who are at high risk of having Huntington's disease overestimate their probability of 
being healthy and make retirement decisions as if they were healthy. @Hoffman2020 show that workers 
overestimate the quality of their match to their 
current employment and are unlikely to look for other opportunities. @Camerer1999 find that 
entrepreneurs are overconfident about the quality of their enterprise, 
which leads to excessive entry and early exit from markets.

In all of these examples,
holding an incorrect belief about a fundamental leads to suboptimal choices with 
potentially high costs. In spite of the abundance of evidence, the scope of the existing research 
is limited in terms of the frameworks it considers. Most of the experimental evidence 
documenting the bias is collected in settings where subjects are passive learners.\footnote{
\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.} They 
observe a noisy signal and report their beliefs over subsequent rounds.\footnote{\citet{Bracha2012} and 
\citet{Mobius2022} are some examples.} Although these studies provide important insights, they are 
not flexible enough 
to incorporate the richer theories that have been proposed more recently. In particular, they 
do not allow the study of endogenous learning or for learning about multiple parameters at once.\footnote{\citet{Coutts2020} 
studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

@@

When the setting features learning traps, even an agent who incorporates all information correctly, 
may fall into learning traps as outlined by @Hestermann2021. These traps are characterized by a 
combination of an incorrect belief and an
optimal action which produce information that confirms the incorrect belief. 
Once an agent falls into a trap, the belief will be stable and even with 
a correctly specified model of learning, they will not be able to abandon their misspecified beliefs. 
If the agent is 
dogmatic about their initial belief, @Heidhues2018 show that they will inevitably fall into a trap
and thus will be able to rationalize and sustain their initial misspecification belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

@Ba2023 moves away from dogmatism and endows the agents with a mechanism through which 
they can abandon incorrect beliefs. This allows them to avoid falling into learning traps. 
To do so agents perform Bayesian hypothesis tests that evaluate which is the more 
likely parameter out of two possibilities. By doing this, she characterizes the set of situations 
in which, even agents who consider alternative paradigms, may become trapped.\footnote{A 
similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has been widely studied in 
behavioral science.\footnote{See \citet{kelley1980} for a review.}
Agents who suffer from misattribution bias will attribute successes to their own ability---the 
ego-relevant parameter---and failures to bad luck---the state. 
Under this model of learning, even an agent who initially has a correct initial 
belief may become overconfident if they observe a sequence of successes. In this case, 
the main driver of the bias is not an initial misspecification or the presence of learning traps, 
it is the updating procedure itself.\footnote{A more general framework that can be used to model 
this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified framework that can be directly 
implemented in a laboratory experiment. In the experiment subjects make choices and receive feedback that depends on their own ability, an exogenous parameter and the choice they made. I track their choices as well as their beliefs about their own ability. The goal is to 
identify which of the 3 forces---the presence of learning traps, misspecified initial beliefs, or misatribution 
bias---better explains the observed behavior. To determine the fit of the models, I compare the behavior
predicted by each theory to the benchmark given by the fully Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the ego-relevance of the parameters 
or if it is a more general phenomenon. I vary the degree of ego-relevance by introducing a 
treatment in which subjects learn about the ability of another participant.
In this treatment, the participants know only the gender and nationality of the other, and thus can 
induce stereotypes---a different type of misspecification.  
If correct learning about the parameters happens at higher rates in the stereotype treatment, 
it would suggest that the 
bias is intrinsically linked to the ego-relevance of the parameter. In contrast, if similar 
biased behavior arises in both treatments, it is more likely that the main driver of these types 
of misspecified beliefs is the updating procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find that the behavior of most 
subjects is better explained by misattribution bias: good news are treated as signaling high ability, while bad news are attributed to a low state. I also find that misattribution is 
no more prevalent in the ego-relevant than in the stereotype condition. This suggests 
that the main driver of the misspecification is the updating procedure; however, the underlying 
mechanism by which the bias is generated may be 
different in both treatments: While in the ego-relevant condition subjects prefer to hold themselves in high esteem, in the stereotype condition
updating seems to be driven by some sort of bias overcorrection---when subjects realize that they underestimated 
the ability of another participant based on their gender and nationality, they compensate by
overestimating their ability.

Finally, I estimate the structural parameters of the models to study model-heterogeneity in the sample.
I find that even at an individual level the behavior is better explained by a general model of 
misattribution bias for most
subjects. There is a smaller group of subjects that can be better explained through dogmatic beliefs
and hypothesis testing and none of them behave in line with the fully Bayesian benchmark.

In what follows I first discuss the theoretical framework and the predictions of each of the theories.
Then I introduce a unifying example and my hypotheses. In section 4 I describe the experimental
design and in section 5 I present the data and the results. Section 6 outlines the estimation of the parameters
and the model fit analysis.

@@

The theories that I consider make predictions in two distinct frameworks. I first detail each theory 
within their original framework and then develop a unifying example that allows me to
compare the predictions of all of them. In framework 1 I consider a setting in which the an agent observes
a continuous output and infers the underlying parameters. In framework 2, the agent observes a binary 
outcome and infers the underlying probability of success, which is related to the parameters. In both frameworks, I consider two nested theories of learning.

@@

An agent is of type $\theta \in \Theta$ and faces an unknown exogenous state $\omega$ drawn from some 
density $f$ over $\Omega$. The agent knows the 
distribution of $\omega$ but not its realized value. His belief about the state, 
$p_0(\omega)$, coincide with the true distribution, $f$; and his belief about the type is $p_0(\theta)$. 
Let the agent's true type be $\theta^{*}$,  and the realized state be $\omega^{*}$.

@@

The agent chooses an action $a\in A$ and observes a noisy outcome $h$. 
The outcome is a function of the agent's type, the state, and
the action. In particular $h = h(\theta^*, \omega^*, a) + \varepsilon$ with $h(\cdot)$ increasing in both $\theta^*$ and $\omega^*$, 
and such that conditional on a pair of parameters $(\theta, \omega)$, there is a unique optimal action. 
Additionally, $\varepsilon\sim N(0, \sigma)$ is noise in the output.

After observing the outcome, the agent updates his beliefs about $\theta$ and $\omega$ using some algorithm and moves on to the next period.
He repeats this process infinitely many times. I make the simplifying assumption that the agent is myopic and chooses the action 
that maximizes the payoff in each period. This assumption simplifies the analysis and plays a role in whether an agent who updates their beliefs using Bayes rule would learn the truth or not, however, for the main theory discussed in this section, the results hold even when relaxing this assumption.

A key notion in this setting is that of a self-defeating equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in \citet{Esponda2016} to this setting with only one agent}. 
A \emph{self-defeating equilibrium} is a belief and action pair, $(p, a)$, such that the agent's belief about their type is misspecified, and the outcome generated by the action is consistent with the misspecified belief. The agent's belief is said to be \emph{stable} when this happens and the average output is

$$\bar{h}(a^*) = h(\hat{\theta}, \hat{\omega}, a)$$
Where $\hat{\theta}$ and $\hat{\omega}$ is the type-state pair on which the misspecified belief places all probability. That is, the agents belief is $p(\hat{\theta}, \hat{\omega}) = 1$ and $p(\theta, \omega) = 0$ for all other values of the parameter.

Within this framework, I consider two nested theories of belief updating: the first one is a dogmatic modeler adapted from @Heidhues2018,The second one is a switcher, as in @Ba2023. The dogmatic modeler can be seen as a switcher with an infinitely sticky initial belief so the two theories are nested. However, outside of the limiting case, they produce different 
predictions about the agent's behavior. I discuss each of them separately in what follows.

@@

A dogmatic agent is characterized by a learning model where he does not update their beliefs about $\theta$, instead, he holds a degenerate belief that places all probability on $\hat{\theta}$, which is potentially misspecified. 
In this case, no matter how much information he gathers against being of type $\hat{\theta}$, he will not update his beliefs about it. 
Any discrepancies between the observed outcomes and his believed type are incorporated using the Bayes rule to update his beliefs about $\omega$. @Heidhues2018 show that, under certain assumptions on the per-period utility,\footnote{The assumptions are that $u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, (ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic modeler will inevitably fall into a self-defeating equilibrium. The equilibrium will be such that the outcomes they observe reinforce their belief on $\omega$ in such a way that as $t\to\infty$ the agent will be sure that the state is some $\omega^{'}$ consistent with their believed type and the observed data. In other words, they will be in a self-defeating equilibrium with a stable belief that places probability one on the incorrect parameters $(\hat{\theta}, \omega_{\infty})$

The mechanism by which the dogmatic agent falls into the self-defeating equilibrium is the following: Suppose the agent holds the misspecified belief that they are type $\hat{\theta}>\theta^*$. For any prior over $\omega$, the agent will be disappointed by the outcome. That is: 
$$\bar{h}(\theta^*, \omega^*, a_t^*) := \frac{1}{T}\sum_t \left[h(\theta^*, \omega^*, a_t^*)+\varepsilon_t\right]<\frac{1}{T}\sum_t \left[h(\hat{\theta}, \mathbb{E}(\omega), a_t^*)+\varepsilon_t\right]$$
Where the right-hand-side of the inequality represents the average outcome that the agent expected.

There are two possible sources for the disappointment: the first is that the realized state is lower than the expected state; the second source is that the agent is of type $\theta^*$ and therefore, for all possible states, his gain will be lower than what he expected.

Because the agent is dogmatic, he will not update his beliefs about $\theta$ and as a consequence will attribute the disappointment to the state being lower than expected. He will continue to update in this way until he converges to a belief about $\omega$ that is stable. Such a belief will explain the observed utility perfectly and allow the agent to rationalize his dogmatic belief about $\theta$. Under the assumptions of @Heidhues2018, there is a unique value of $\omega$ at which the belief is stable, I will refer to such value as $\omega_\infty$. This mechanism is illustrated in Example 1.

**Example 1: ** 
_Set $A = \Omega=[\underline{a}, \infty)$ and consider a student with intrinsic ability $\theta^*\geq 0$ who faces a grading procedure 
$\omega^*$ that is unknown to them. 
The student knows that a higher $\omega^*$ is more likely to yield a higher grade. In particular, she knows the grade is given by 
$(\theta^*+a)\omega^*$._

_The student must choose an effort level $a$, which determines her grade. For whatever the chosen effort level, 
the student must pay a cost $c(a) = \frac{1}{2}a^2$ and observes a grade of $(\theta^*+a)\omega^*$. She repeats 
this process for infinitely many periods. 
Assume also that the student's prior is such that $\mathbb{E}[\omega]= \omega^*$ and she is dogmatic 
about being of 
type $\hat{\theta}>\theta^*$.\footnote{The example is illustrated for an overconfident student but 
the results are symmetric for a dogmatic student who initially 
places probability one on some $\tilde{\theta}<\theta^*$.}
Therefore, the student's payoff in period $t$ is given by _

@@

_Under this specification, the myopic optimal effort level is $a_t^* = \omega^*$. 
Because the student does not know $\omega^*$,
he will choose $a_t = \mathbb{E}_t(\omega)$ where the expectation is taken with respect to the 
student's belief at the beginning of
period $t$. If she does not revise his effort choice for $k$ periods, she will receive 
an average utility of 
$(\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$ but he was expecting an average utility of 
$(\hat{\theta}+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$. In response, she will apply Bayes 
rule to update his beliefs about
$\omega$ to get the posterior belief with 
$\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}$
which is lower than the initial belief. This will cause the student to choose a lower effort at $t+k$. As a result, she will again
receive an average utility that is lower than what he expected which will cause her belief to drift further down. This process will
continue until the average utility equals her expected utility under the dogmatic belief that assigns probability 1 to $\hat{\theta}$. 
At that point, the student will have reached a self-defeating equilibrium and she will continue to choose sub-optimal effort forever. _

@@

In particular, for an agent that starts with a dogmatic belief that they are of type $\hat{\theta}$ but is willing to consider
the alternative explanation that they are of type $\tilde{\theta}$, the agent will switch to the alternative if:

@@

Where $h^t$ is the history of outcomes up to time $t$, and $\alpha$ is the switching threshold.\footnote{Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.} By keeping track of the 
likelihood ratio, the agent can perform a \emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best fits 
the data.\footnote{In a related problem \citet{Schwarstein2021} proposes a similar updating procedure which relies on the Bayesian hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}

By allowing the agent to keep track of the likelihoods and switching to an alternative type, the switcher can avoid the 
self-confirming equilibria. However, if the prior belief on $omega$ is sufficiently tight around
a self-defeating equilibrium, the switcher might look identical to the dogmatic even in a case where $\alpha$ is not too large.
This happens because under the agent's prior, the likelihood ratio is unlikely to grow as fast as it is needed to escape 
the self-defeating equilibrium. In such situations, we say that the misspecified belief is persistent.

@@

As in framework 1, the agent is of some type $\theta^* \in \Theta$ and the state is $\omega \sim F(\Omega)$. In this case, the agent chooses an action $a\in A$ and observes a binary outcome that is either a success or a failure. Denote the outcome by $o \in \{s,f\}$. The probability of observing a success is increasing in $\theta$ and in $\omega$. Whenever the agent observes a success, he gets a payoff $v>0$ and whenever the outcome is a failure, the payoff is 0.  In addition, the probability of success is such that for each state, there is a unique optimal action that maximizes the agent's expected payoff. Therefore, the probability of success can be seen as a monotone transformation of the utility from Framework 1.

I focus on two nested theories that have been widely studied within this framework: Full Bayesian updating and self-serving attribution bias. I explain each of these classical models of belief updating in what follows

@@

A Self-Serving Bayesian is an agent who uses a biased version of Bayes rule to update their beliefs. They will update their beliefs about the state $\omega$ and his type $\theta$ simultaneously by over-attributing successes to a high value of $\theta$ and under-estimating the role of higher $\omega$. Similarly, he will attribute failure to a low state to a greater degree than an unbiased agent would. This sort of biased updating is referred to as \emph{misattribution bias}.

>>>>

---
title: |
  | Learning with Misspecified Models: The Case of Overestimation
  | \href{https://jimena-galindo.github.io/fileRepo/draft.pdf}{\normalsize Click Here for the Latest Version}

date: '`r format(Sys.time(), "%B %d, %Y")`'
author: "Jimena Galindo"
abstract: "I design a framework and a laboratory experiment that allow for the comparison of multiple theories of misspecified learning. I focus on a framework with endogenous information and a data-generating process ruled by two fundamentals: an ego-relevant parameter and a state. Within this framework I study three forces that can lead to misspecified beliefs: initial misspecifications, learning traps and biased updating. I find that biased updating is the main driver of misspecified beliefs in the lab. In addition, I vary the degree of ego-relevance of the parameter by introducing a stereotype treatment. The data is consistent with biased updating in both cases but for potentially different reasons: when learning about themselves, subjects attribute successes to their own ability and failures to luck. Instead, in the stereotype treatment, they compensate for initial negative biases by over-attributing positive signals to the ability of others. This translates into similar observed choices but different dynamics in beliefs."

@@

bibliography: references.bib
fontsize: 12pt 
geometry: margin=2.5cm
link-citations: true
linkcolor: mypink
filecolor: myblue
citecolor: mypink
urlcolor: mypink

@@

As an example, consider a student who needs to decide how much effort to put into studying for an exam. Their decision will depend on two factors: their belief about their intrinsic ability and their belief about how difficult the exam will be. The outcome they observe will be affected by how much they decide to study. Imagine that the student exerts a moderate amount of effort and gets a surprisingly good grade. Did they get a good grade because they are smarter than they thought? Or because the exam was easier than they had anticipated? Their future exam-preparation strategy will depend fundamentally on which line of reasoning they take. This feedback loop is referred to as an \emph{endogenous information process} and is at the center of the forces I study.

To understand what main forces are at play, I compare a set of theories that model learning in settings with 
endogenous information and can rationalize the persistence of misspecified beliefs. I develop a unifying framework that nests multiple theories of learning and generates testable predictions for each of them. Then, I cast this framework in a laboratory experiment and test the predictions to identify which of the theories are consistent with the behavior I observe in the lab. The experiment features an agent who needs to learn two parameters: one that pertains to their own characteristics (an \emph{ego-relevant} parameter), and an exogenously determined state. In the context of the student preparing for exams, the two parameters  would be their intrinsic ability and the difficulty of the exam.

Misspecified beliefs about an ego-relevant parameter are often referred to as overconfidence (or underconfidence) and have been documented by behavioral scientists\footnote{\citet{kelley1980} provides a review of the psychology literature.} and economists\footnote{See \citet{benjamin2019} for a review of the literature in economics.} in a variety of settings.  @Oster2013 find subjects who are at high risk of having Huntington's disease overestimate their probability of being healthy, and they make retirement decisions as if they were healthy. @Hoffman2020 show workers overestimate the quality of their match to their current employment and are unlikely to look for other opportunities. @Camerer1999 find entrepreneurs are overconfident about the quality of their enterprise, which leads to excessive entry and early exit from markets.

In all of these examples, holding an incorrect belief about a fundamental leads to suboptimal choices with potentially high costs. In spite of the abundance of evidence, the scope of the existing research is limited in terms of the frameworks it considers. Most of the experimental evidence documenting the bias is collected in settings where subjects are passive learners.\footnote{\citet{Gotte2022} and \citet{Ozyilmaz2022} are exceptions that study settings with endogenous information processes.} They observe a noisy signal and report their beliefs over subsequent rounds.\footnote{\citet{Bracha2012} and \citet{Mobius2022} are some examples.} Although these studies provide important insights, they are not flexible enough to incorporate the richer theories that have been proposed more recently. In particular, they do not allow the study of endogenous learning or for learning about multiple parameters at once.\footnote{\citet{Coutts2020} studies an environment with an ego-relevant parameter and an exogenous state but does not 
incorporate the endogenous information process.}

@@

When the setting features learning traps, even an agent who incorporates all information correctly may fall into learning traps, as outlined by @Hestermann2021. These traps are characterized by a combination of an incorrect belief and an optimal action that produce information that confirms the incorrect belief. Once an agent falls into a trap, the belief will be stable and even with a correctly specified model of learning, they will not be able to abandon their misspecified beliefs. If the agent is dogmatic about their initial belief, @Heidhues2018 show they will inevitably fall into a trap and thus will be able to rationalize and sustain their initial misspecification belief.\footnote{\citet{Gotte2022} study the case of agents with dogmatic initial beliefs in a 
laboratory experiment.}

@Ba2023 moves away from dogmatism and endows the agents with a mechanism through which they can abandon incorrect beliefs. This allows them to avoid falling into learning traps. To do so, agents perform Bayesian hypothesis tests that evaluate which of two possible parameters is more likely. By doing so, she characterizes the set of situations in which even agents who consider alternative paradigms may become trapped.\footnote{A similar mechanism is proposed by \citet{Schwarstein2021} in a setting with persuasion.}

Lastly, misattribution bias is the more classical explanation and has been widely studied in behavioral science.\footnote{See \citet{kelley1980} for a review.} Agents who suffer from misattribution bias will attribute successes to their own ability---the 
ego-relevant parameter---and their failures to bad luck---the state. Under this model of learning, even an agent who initially has a correct initial belief may become overconfident if they observe a sequence of successes. In this case, the main driver of the bias is not an initial misspecification or the presence of learning traps; it is the updating procedure itself.\footnote{A more general framework that can be used to model this bias has also been proposed by \citet{Brunnermeier2005} and empirically studied by \citet{Bracha2012}.}

These theories provide the main building blocks for a simplified framework that can be directly implemented in a laboratory experiment. In the experiment subjects make choices and receive feedback that depends on their own ability, an exogenous parameter and the choice they made. I track their choices as well as their beliefs about their own ability. The goal is to identify which of the three forces---the presence of learning traps, misspecified initial beliefs, or misatribution bias---better explains the observed behavior. To determine the fit of the models, I compare the behavior predicted by each theory with the benchmark given by the fully-Bayesian updating procedure.

I also study whether the learning mechanism is inherently linked to the ego-relevance of the parameters or if it is a more general phenomenon. I vary the degree of ego-relevance by introducing a treatment in which subjects learn about the ability of another participant. In this treatment, the participants know only the gender and nationality of the other, and thus can induce stereotypes---a different type of misspecification. If correct learning about the parameters happens at higher rates in the stereotype treatment, this finding would suggest the bias is intrinsically linked to the ego-relevance of the parameter. By contrast, if similar biased behavior arises in both treatments,  the main driver of these types of misspecified beliefs is more likely to be the updating procedure itself or the endogenous information process.

Although some agents do fall into learning traps, I find the behavior of most subjects is better explained by misattribution bias: good news are treated as signaling high ability, whereas bad news are attributed to a low state. I also find misattribution is no more prevalent in the ego-relevant than in the stereotype condition. This finding suggests the main driver of the misspecification is the updating procedure; however, the underlying mechanism by which the bias is generated may be different in both treatments. Whereas subjects in the ego-relevant condition prefer to hold themselves in high esteem, in the stereotype condition, updating seems to be driven by some sort of bias overcorrection---when subjects realize they underestimated 
the ability of another participant based on their gender and nationality, they compensate by overestimating their ability.

Finally, I estimate the structural parameters of the models to study model-heterogeneity in the sample. I find that even at an individual level, the behavior is better explained by an average model of misattribution bias for most subjects. A smaller group of subjects exists that can be better explained through dogmatic beliefs and hypothesis testing, and very few of them behave in line with the fully Bayesian benchmark.

In what follows, I first discuss the theoretical framework and the predictions of each of the theories. Then, I introduce a unifying example and my hypotheses. In section 4, I describe the experimental design, and in section 5, I present the data and the results. Section 6 outlines the estimation of the parameters and the model fit analysis.

@@

The theories I consider make predictions in two distinct frameworks. I first detail each theory 
within their original framework and then develop a unifying example that allows me to
compare the predictions of all of them. In framework 1, I consider a setting in which an agent observes
a continuous output and infers the underlying parameters. In framework 2, the agent observes a binary 
outcome and infers the underlying probability of success, which is related to the parameters.

@@

An agent of type $\theta \in \Theta$ faces an unknown exogenous state $\omega$ drawn from 
density $f$ over $\Omega$. The agent knows the 
distribution of $\omega$ but not its realized value. His belief about the state $\omega\in\Omega$ is 
$p_0(\omega)$ and coincides with the true distribution, $f$. His belief about the type 
$\theta\in\Theta$ is $p_0(\theta)$. 
Let the agent's true type be $\theta^{*}$,  and the realized state be $\omega^{*}$.

@@

The agent chooses an action $a\in A$ and observes a noisy outcome $h$. 
The outcome is a function of the agent's type, the state, and
the action. In particular, $h = h(\theta^*, \omega^*, a) + \varepsilon$ with $h(\cdot)$ increasing in both $\theta^*$ and $\omega^*$, 
and such that conditional on a pair of parameters $(\theta, \omega)$, there is a unique optimal action. 
Additionally, $\varepsilon\sim N(0, \sigma)$ is noise in the outcome.

After observing the outcome, the agent updates his beliefs about $\theta$ and $\omega$ using some algorithm and moves on to the next period. He repeats this process infinitely many times. I make the simplifying assumption that the agent is myopic and chooses the action that maximizes the payoff in each period. This assumption simplifies the analysis and plays a role in whether an agent who updates his beliefs using Bayes rule would learn the truth or not, however, for the main theory discussed in this section, the results hold even when relaxing this assumption as shown in @Heidhues2018.

A key notion in this setting is that of a self-defeating equilibrium\footnote{This notion is an adaptation of the Berk-Nash equilibrium in \citet{Esponda2016} to this setting with only one agent}. 
A \emph{self-defeating equilibrium} is a belief and action pair, $(p, a^*)$, such that the agent's belief about his type is misspecified, and the outcome generated by the action is consistent with the misspecified belief. The agent's belief is said to be \emph{stable} when this happens.

Given the choice that sustains the self-defeating equilibrium, they observe the \emph{average output} given by
$$\bar{h}(a^*) =\frac{1}{T}\sum_t \left[h(\hat{\theta}, \hat{\omega}, a^*)+\varepsilon_t\right]$$ 
where $\hat{\theta}$ and $\hat{\omega}$ is the type-state pair on which the misspecified belief places all probability. That is, the agents belief is $p(\hat{\theta}, \hat{\omega}) = 1$ and $p(\theta, \omega) = 0$ for all other values of the parameter.

Within this framework, I consider two nested theories of belief updating: the first one is a dogmatic modeler adapted from @Heidhues2018. The second one is a switcher, as in @Ba2023. The dogmatic modeler can be seen as a switcher with an infinitely sticky initial belief so the two theories are nested. However, outside of the limiting case, they produce different 
predictions about the agent's behavior. I discuss each of them separately in what follows.

@@

A dogmatic agent is characterized by a learning model where he does not update his beliefs about $\theta$ and instead, holds a degenerate belief that places all probability on $\hat{\theta}$, which is potentially misspecified. 
In this case, no matter how much evidence he gathers against being of type $\hat{\theta}$, he will not update the beliefs about his type. 
Any discrepancies between the observed outcomes and his believed type are incorporated using the Bayes rule to update his beliefs about $\omega$. @Heidhues2018 show that, under certain assumptions on the per-period utility,\footnote{The assumptions are that $u$ is twice continuously differentiable with: (i)$u_{aa}<0$ and $u_a(\underline{a},  \theta, \omega)>0>u_a(\Bar{a}, \theta, \omega)$, (ii) $u_{\theta}, u_{\omega}>0$ and (iii) $u_{a\theta}<0$ and $u_{a\omega}>0$. Where $\underline{a}$ is the
minimal action and $\Bar{a}$ is the maximal one. The direction of the derivatives is a normalization
and the results would hold even when the signs are reversed.} a dogmatic modeler will inevitably fall into a self-defeating equilibrium. The equilibrium will be such that the outcomes he observes reinforce his belief on $\omega$ in such a way that as $t\to\infty$ the agent will be sure that the state is some $\omega^{'}$ consistent with their believed type and the observed data. In other words, he will be in a self-defeating equilibrium with a stable belief that places probability one on the incorrect parameters $(\hat{\theta}, \omega_{\infty})$

The mechanism by which the dogmatic agent falls into the self-defeating equilibrium is the following: suppose the agent holds the dogmatic and misspecified belief that they are type $\hat{\theta}>\theta^*$. For any prior over $\omega$, the agent will be disappointed by the outcome. That is: 
$$\frac{1}{T}\sum_t \left[{h}(\theta^*, \omega^*, a_t^*)+\varepsilon_t\right] <\mathbb{E_w} \left[h(\hat{\theta}, \omega, a_t^*)\right]$$
where the right-hand-side of the inequality represents the average outcome that the agent expected given his beliefs about their type and the state. The left hand side is what they observed.

There are two possible sources for the disappointment: the first is that the realized state $\omega^*$ is lower than what the agent expected, $\mathbb{E}[\omega]$; the second source is that the agent is of type $\theta^*$ and therefore, for all possible states, his gain will be lower than what he expected when the variance of the error term is low or $T$ is large. That is, when $\frac{1}{T}\sum_t\varepsilon_t$ is close to zero.

Because the agent is dogmatic, he will not update his beliefs about $\theta$ and, as a consequence, will attribute the disappointment to the state being lower than expected. He will continue to update in this way until he converges to a belief about $\omega$ that is stable. Such a belief will explain the observed utility perfectly and allow the agent to rationalize his dogmatic belief about $\theta$. Under the assumptions of @Heidhues2018, there is a unique value of $\omega$ at which the belief is stable, I will refer to such value as $\omega_\infty$. This mechanism is illustrated in Example 1.

**Example 1: ** 
_Set $A = \Omega=[\underline{a}, \infty)$ and consider a student (she) with intrinsic ability $\theta^*\geq 0$ who faces a grading procedure $\omega^*$ that is unknown to her. The student knows that a higher $\omega^*$ is more likely to yield a higher grade. In particular, she knows the grade is given by $(\theta^*+a)\omega^*$._

_The student must choose an effort level $a$, which determines her grade. For whatever the chosen effort level, the student must pay a cost $c(a) = \frac{1}{2}a^2$. She repeats this process for infinitely many periods. Assume also that the student's prior is such that $\mathbb{E}[\omega]= \omega^*$ and she is dogmatic about being of type $\hat{\theta}>\theta^*$.\footnote{The example is illustrated for an overconfident student but the results are symmetric for a dogmatic student who initially places probability one on some $\tilde{\theta}<\theta^*$.} Therefore, the student's payoff in period $t$ is given by _

@@

_Under this specification, the myopic optimal effort level is $a_t^* = \omega^*$. Because the student does not know $\omega^*$,
she will choose $a_t = \mathbb{E}_t(\omega)$ where the expectation is taken with respect to the student's belief at the beginning of
period $t$. If she does not revise his effort choice for $k$ periods, she will receive an average utility of $(\theta^*+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$ but was expecting an average utility of $(\hat{\theta}+a_t^*)\omega^* - \frac{1}{2}a_t^{*2}$. In response, she will apply Bayes rule to update his beliefs about $\omega$ to get the posterior belief with $\mathbb{E}_{t+k}[\omega] = \frac{(\theta^{*} + \omega^{*})\omega^{*}}{\hat{\theta} + \omega^{*}}$ which is lower than the initial belief. This will cause the student to choose a lower effort at $t+k$. As a result, she will again receive an average utility that is lower than what he expected which will cause her belief to drift further down. This process will continue until the average utility equals her expected utility under the dogmatic belief that assigns probability 1 to $\hat{\theta}$. At that point, the student will have reached a self-defeating equilibrium and she will continue to choose sub-optimal effort forever._

@@

In particular, for an agent that starts with a dogmatic belief that they are of type $\hat{\theta}$ but is willing to consider
the alternative explanation that they are of type $\tilde{\theta}$, the agent will switch to the alternative if

@@

where $h^t$ is the history of outcomes up to time $t$ and $\alpha$ is the switching threshold.\footnote{Notice that if $\alpha \to \infty$, the behavior of the switcher will be indistinguishable from that of the Dogmatic modeler. 
In this sense, the switcher is a generalization of the dogmatic type.} By keeping track of the 
likelihood ratio, the agent can perform a \emph{Bayesian hypothesis test} and adopt the Dogmatic belief that best fits 
the data.\footnote{In a related problem \citet{Schwarstein2021} proposes a similar updating procedure which relies on the Bayesian hypothesis test. However, in their model there is a sender who optimally chooses to propose a model that fits the data}

By allowing the agent to keep track of the likelihoods and switching to an alternative type, the switcher can avoid the 
self-confirming equilibria. However, if the prior belief on $\omega$ is sufficiently tight around
a self-defeating equilibrium, the switcher might look identical to the dogmatic even in a case where $\alpha$ is not too large.
This happens because under the agent's prior, the likelihood ratio is unlikely to grow as fast as it is needed to escape 
the self-defeating equilibrium. In such situations, we say that the misspecified belief about $\theta$ is \emph{persistent}.

@@

As in framework 1, the agent is of some type $\theta^* \in \Theta$ and the state is $\omega$ is distributed according to  $F(\Omega)$. In this case, the agent chooses an action $a\in A$ and observes a binary outcome that is either a success or a failure. Denote the outcome by $o \in \{s,f\}$. The probability of observing a success is increasing in $\theta$ and in $\omega$. Whenever the agent observes a success, he gets a payoff $v>0$ and whenever the outcome is a failure, the payoff is 0.  In addition, the probability of success is such that for each state, there is a unique optimal action that maximizes the agent's expected payoff. Therefore, the probability of success can be seen as an order-preserving transformation of the utility from Framework 1.

I focus on two nested theories that have been widely studied within this framework: full Bayesian updating and self-serving attribution bias. I explain each of these classical models of belief updating in what follows.

@@

A Self-Serving Bayesian is an agent who uses a biased version of Bayes rule to update his beliefs. He will update his beliefs about the state $\omega$ and his type $\theta$ simultaneously by over-attributing successes to a high value of $\theta$ and under-estimating the role of higher $\omega$. Similarly, he will attribute failure to a low state to a greater degree than an unbiased agent would. For instance, if the student from the example attributed a high grade to being of high ability while attributing low grades to a harsh grading system. This sort of biased updating is referred to as \emph{misattribution bias}.
